Stop Lying to Me: New Visual Tools to Choose the Most Honest Nonlinear Dimension Reduction

### 2025-06-26 paper.qmd
Changed PBMC3k text
Updated Figure 15

@fig-pbmc-rmse shows RMSE across a range of binwidths ($a_1$) for each of the layouts in @fig-NLDR-variety. The layouts were made using tSNE, UMAP, PHATE, PaCMAP, and TriMAP with various hyper-parameter settings. Lines are color coded to match the color of the layouts shown on the right. Lower RMSE indicates the better fit. Using a range of binwidths shows how the model changes, with possibly the best model being one that is universally low RMSE across all binwidths. It can be seen that layout f is sub-optimal with universally higher RMSE. Layout a, the published one, is better but it is not as good as layouts b, d, or e. With some imagination layout d perhaps shows three barely distinguishable clusters. Layout e shows three, possibly four, clusters that are more separated. The choice reduces from eight to these two. Layout d has slightly better RMSE when the $a_1$ is small, but layout e beats it at larger values. Thus we could argue that layout e is the most accurate representation of the cluster structure, of these eight.

@@

<!--mnist/04_gen_mse_for_diff_methods.R-->
```{r}
#| label: combine-data-mnist

@@

<!-- Fit the model and compute error-->
<!--mnist/05_gen_model_with_tSNE.R-->
```{r}
#| label: hexbin-mnist

@@

<!--mnist/03_gen_images.R-->
```{r}
#| label: img-diff-pos

@@

right_top_img <- plot_digit_img(
  img_right_top, palette = "Blues 2", title_text = "b")

@@

middle_img <- plot_digit_img(
  img_middle, palette = "Greens 3", title_text = "c")

@@

right_bottom_img <- plot_digit_img(
  img_right_bottom, palette = "Purples 2", title_text = "d")

@@

imge_error_sample_outside <- plot_digit_img(
  img_error_outside, palette = "Grays", title_text = "e")

@@

imge_error_sample_inside <- plot_digit_img(
  img_error_inside, palette = "Oranges", title_text = "f")

>>>>

@fig-pbmc-rmse shows RMSE across a range of binwidths ($a_1$) for each of the layouts in @fig-NLDR-variety. The layouts were generated using tSNE and UMAP with various hyper-parameter settings, while PHATE, PaCMAP, and TriMAP were applied using their default settings. Lines are color coded to match the color of the layouts shown on the right. Lower RMSE indicates the better fit. Using a range of binwidths shows how the model changes, with possibly the best model being one that is universally low RMSE across all binwidths. It can be seen that layout f is sub-optimal with universally higher RMSE. Layout a, the published one, is better but it is not as good as layouts b, d, or e. With some imagination layout d perhaps shows three barely distinguishable clusters. Layout e shows three, possibly four, clusters that are more separated. The choice reduces from eight to these two. Layout d has slightly better RMSE when the $a_1$ is small, but layout e beats it at larger values. Thus we could argue that layout e is the most accurate representation of the cluster structure, of these eight.

@@

<!--mnist/03_gen_mse_for_diff_methods.R-->
```{r}
#| label: combine-data-mnist

@@

<!-- Fit the model and compute error-->
<!--mnist/04_gen_model_with_tSNE.R-->
```{r}
#| label: hexbin-mnist

@@

```{r}
#| label: img-diff-pos

@@

right_top_img <- plot_digit_img(
  img_right_top, color_on = "#375cb1", title_text = "b")

@@

middle_img <- plot_digit_img(
  img_middle, color_on = "#008000", title_text = "c")

@@

right_bottom_img <- plot_digit_img(
  img_right_bottom, color_on = "#4c389a", title_text = "d")

@@

imge_error_sample_outside <- plot_digit_img(
  img_error_outside, color_on = "#000000", title_text = "e")

@@

imge_error_sample_inside <- plot_digit_img(
  img_error_inside, color_on = "#ff7f00", title_text = "f")
